Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-10 paper-new/concrete-term-model.tex
discussion of adequacy for term semantics

\max{TODO: Eric needs to add a discussion of adequacy/clock stuff here}

>>>>

% \max{TODO: Eric needs to add a discussion of adequacy/clock stuff here}
\eric{Check this}

We now define from the above term model a ``big-step'' semantics. More
concretely, our goal is to define a function 
$-\Downarrow : \{M \,|\, \cdot \vdash M : \nat \} \rightharpoonup \mathbb{N} + {\mho}$.

To begin, we must first discuss another aspect of guarded type theory,
\emph{clocks}, originally introduced in \cite{atkey-mcbride2013}. The
constructions of guarded type theory ($\later$, $\nxt$, and $\fix$) we have been
using may be indexed by objects called \emph{clocks}. A clock intuitively serves
as a reference relative to which steps are counted. For instance, given a clock
$k$ and type $X$, the type $\later^k X$ represents a value of type $X$ one unit
of time in the future according to clock $k$.
%
Up to this point, all uses of guarded constructs have been indexed by a single
clock $k$. In particular, we have for each clock $k$ a type $\li^k X$. The
notion of \emph{clock quantification} allows us to pass from the guarded setting
to the usual set-theoretic world. Clock quantification enables us to encode
coinductive types using guarded recursion. That is, given a type $A$ with a free
variable $k : \Clock$, we can form the type $A^{gl} := \forall k. A$. Given a
functor $F$ that satisfies a technical condition, if $A$ is a guarded recursive
type satisfying $A \cong F(\later^k A)$, then the type $A^{gl}$ has a final
$F$-coalgebra structure (see Theorem 4.3 of
\cite{kristensen-mogelberg-vezzosi2022}).

Consider the functor $F(X) = \mathbb{N} + {\mho} + X$. By the definition of the
guarded lift, we have that
% 
$\li^k \mathbb{N} 
  \cong \mathbb{N} + {\mho} + \later^k (\li \mathbb{N}) 
  \cong F(\later \li^k \mathbb{N})$.
%
Furthermore, it can be shown that the functor $F$ satisfies the technical
condition in the statement of the above-mentioned theorem. Thus, the theorem
implies that the type $\li^{gl} \mathbb{N} := \forall k. \li^k \mathbb{N}$ is a
final $F$-coalgebra.
%
Given a type $Y$, Capretta's \emph{delay monad} \cite{lmcs:2265},
which we write as $\text{Delay}(Y)$, is defined as the coinductive type
generated by $\tnow : Y \to \delay(Y)$ and $\tlater : \delay(Y) \to \delay(X)$.
We have that $\delay (\mathbb{N} + {\mho})$ is a final coalgebra for the functor
$F$ defined above. This means $\li^{gl} \mathbb{N}$ is isomorphic to
$\delay(\mathbb{N} + {\mho})$.

We can define a notion of termination in $i$ steps for the Delay monad $d \da^i
n_?$ where $d : \delay(\mathbb{N} + {\mho})$ and $n_? : \mathbb{N} + {\mho}$. From
this, can define a partial function $\delay (\mathbb{N} + {\mho})
\rightharpoonup \mathbb{N} + {\mho}$ by existentially quantifying over $i$
for which $d \da^i n_?$.

Now consider the function $\{M \,|\, \cdot \vdash M : \nat \} \to \li^{gl}
\mathbb{N}$ defined by $\lambda M. \lambda k .\sem{M}$. Composing this with the
above partial function yields the desired partial function $-\Downarrow : \{M
\,|\, \cdot \vdash M : \nat \} \rightharpoonup \mathbb{N} + {\mho}$. It is
then straightforward to see that this big-step semantics respects the equational
theory and that the denotations of the diverging term $\Omega$, the error term
$\mho$, and a syntactic natural number value $n$ are all distinct.
